---
title: "Replicating an Analysis on Hurricane Disturbance and Stream Water Chemistry in Luquillo, Puerto Rico"
author: "Ethan Mathews"
date: "2026-09-27"
format:
  html:
    toc: true 
    toc-depth: 5
---

# Background

The study examines how Hurricane Hugo affected stream water chemistry in eight forested watersheds in Puerto Rico. In this analysis we examine the Bisley Quebrada 1, Bisley Quebrada 2, Bisley Quebrada 3, and Puente Roto Mameyes over a 9-week average. Researchers compared water chemistry before and after the hurricane and found that nitrate, potassium, and ammonium increased significantly for up to two years. These increases occurred because the hurricane damaged vegetation, creating more plant material that decomposed and released nutrients into the streams. Overall, hurricanes can significantly disrupt nutrient cycling in tropical forests.



# Data

```{r}
#| output: false

library(tidyverse)


```

The following data files for this analysis were downloaded from [EDI Data Portal](https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-luq.20.4923064) and can be found in the folder raw_data:

- QuebradaCuenca1-Bisley.csv
- QuebradaCuenca2-Bisley.csv
- QuebradaCuenca3-Bisley.csv
- RioMameyesPuenteRoto.csv

Each .csv file was read in and filtered to the date range from January 5th, 1988 to December 26th, 1994. The relevant columns corresponding to Sample ID, Sample Date, and concentrations for K, NO$_3$, Mg, Ca, and NH$_4$ were also selected.

# Methods

The water chemistry data from four the watersheds (BQ1, BQ2, BQ3, and PRM) were filtered to include samples from 1988–1994. A 9-week moving average was calculated for potassium (K), magnesium (Mg), nitrate (NO3-N), calcium (Ca), and ammonium (NH4-N) concentrations to examine changes in stream chemistry over time. The resulting averages were used to compare water chemistry before and after Hurricane Hugo. 



# Results

```{r}
#| echo: false

fig3 <- read_csv("../output/clean_data.csv")

hurricane_date <- as.Date("1989-09-18")

ggplot(data = fig3, mapping = aes(
    x = window_start,
    y = concentration, 
    linetype = Site
)) + 
  
  geom_line() + 
  
  geom_vline(xintercept = hurricane_date, linetype = "dashed", linewidth = 0.4) +
  
  theme_bw() + 
  
  labs(
    title = "Hurricane effects on stream chemistry",
    x = "Years", 
    y = NULL, 
    linetype = NULL) +
  
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  
  theme(
    strip.placement = "outside",
    strip.background = element_blank(),
    strip.text.y.left = element_text(angle = 90),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()) +
  
  facet_grid(vars(nutrient), scales = "free_y", switch = "y") +
  
  theme(strip.placement = "outside") +
  
  scale_x_date(sec.axis = dup_axis())



```
